Radiocarbon measurement of CO~2~ from carbonate minerals using a gas ion source with an open split interface

### 2022-03-29 doc/AMS-15/AMS-15_paper/AMS-15_paper.Rmd
General reorg and edits.

Measurements are corrected for process and measurement blank using a linear correction that assumes that the corrected fraction modern of blank samples should be zero, samples with the fraction modern of the standard need no correction, and applies a proportional correction to samples of intermediate fraction modern. (@roberts_14c_2019) The correction uses measured values of standard gas and blank samples for correction of contaminant carbon and machine blank. We use equation 1 for this correction.

$$ R_s = R_m - \frac{R_b(R_{ox}-R_m)}{R_{ox}} $$

Where $R_s$ is the ratio of the unknown sample, $R_m$ is the measured ratio, $R_b$ is the ratio of the blank, and $R_{ox}$ is the ratio of the normalizing standard. The errors in the blanks and standards are propagated with errors of the samples to produce the final error estimate for the sample. This correction does not account for the effect of a blank with constant mass or current on samples of varying mass or current. For all measurements used to assess the performance of the method and for the Haiti core comparison, sample, standard, and blank masses were kept constant. Sample current did vary, but we believe the linear correction described above produces the most accurate results within this range. We discuss this constant contaminant blank in the results below. Data reduction is done using a package written in R (@r_core_team_r_2022) which is published on github (<https://github.com/blongworth/hgis>).

@@

All results in supplementary materials

Determination of ideal flows.
Testing is needed to determine minimum sample flow to avoid atmosphere entering the split. TODO: move flow ratio test to results.

## Method development

@@

Initial work with the NEC source established optimal source parameters for consistent source performance. We used tank CO2 connected through a regulator to the ion source to measure source performance over a range of CO2 flows. Best currents for the source are around 10uA 12C- or roughly 1/10 of graphite currents. Best currents were produced at a CO2 flow rate of 7.5uLs-1 CO2, but the source produces near optimal currents with flows from 7.5- XXX ul/min CO2. Adding helium to the gas mixture produced similar currents to pure CO2 at the same CO2 flow rates. This is generally comparable with other hybrid sources @Xu2007, although improved target design and continued optimization of source parameters may improve performance @Salazar2013. Typical currents for this study were between 4-9uA and average current for all samples was `r curmean`. The source typically produces best current at around 7.5uLs-1 CO2. Several studies have shown best currents at lower CO2 flow rates (@Xu2007, @Ruff2010). While efficiency of the source is best at very low flow rates (<1.5ul/min), measuring samples with flow set for optimal current produces more precise results due to higher count rates.

Source currents are variable by day and by target under similar source, inlet, and sample conditions. Capillary condition, source and gas delivery arm alignment, slight variability in target geometry, source condition, and performance of the ionizer and cesium may all be factors. Current also varies by target for a given set of gas and source conditions. Some targets are clearly "bad", producing currents much lower than other targets for a given set of conditions. Capillary clogging can be an issue, especially when sampling from vials with acid.

@@

### Currents

@@

Figure1: Current vs flow

Sample size tests.

Aimed to fill vial initially

No current degradation down to X. Used Y as optimal mass and backfilled with helium.

@@

Setting flow to the source to 10uL/min provides excess CO2 to source until the helium dilution ratio increases to \> 1:1. Another approach may be to backfill vials with helium and run at a higher dilution ratio and flow to the source. This should produce a smaller change in He:CO2 over the course of a run. MICADAS runs at 95% He:CO2, which would mean producing \< 0.5mL CO2 in a 7mL vial and backfilling with helium before displacing to the source. Some intermediate dilution may be appropriate. Alternatively, vials could be flushed with helium prior to acidification, rather than evacuating, leading to some concentration of CO2 in He at slightly \> 1 ATM.

Testing long runs of dilution of CO2 with displacing He and diluting He with CO2 show that the setup is still optimized for low flow to the source, even with larger diameter delivery capillaries. There appears to be a constriction in the split-source tubing that is keeping flows low. Current focus is on the large-sample 7mL method, so this will be diagnosed as time allows.

Targets have shown "pulses" of higher current. Not sure why this is, but I suspect it's either buildup of graphite or other available carbon on the target, or target switching to the "high state" as sometimes happens with solid targets.

First tests with carbonate in vials were hit-or-miss. Problems with capillary clogging and low currents led to poor results for several samples. A pair of NOSAMS2 samples produced good currents, results agreed well, but were about 3% higher than expected Fm. A second set of carbonates produced better currents (\~5uA) and better results. Agreement of replicates varied from 3 to 16 permil. Near-modern NOSAMS2 was very close to the consensus value, while C-2 and C-1 were elevated by 2 and 3.6% respectively, indicating that a blank correction is needed.

Two more recent carbonate tests produced stable currents and backgrounds. Sample, target blank, and sample blank currents were relatively constant across samples, so a large blank correction can be used to correct results. Samples from both runs showed a linear relationship between difference from expected Fm and Fm of the sample, but in one sample set the relationship was positive, and in the other, negative. Since the intercept with 0-offset was close to 0 Fm, I think this is a problem with normalization, rather than blank correction. Standards run at the end of the day are higher than those run at the beginning, which may give clues as to what's going on.

### Contaminant carbon

@@

#### Current dependent blank

@@

$$I_{sample} + I_{blank} = I_{meas}$$
$$Fm_mI_m=Fm_bI_b+Fm_sI_s$$

@@

0.1 μA blank current with 10 μA total = 1% blank
LBC vs constant contaminant/MBC.

#### Current independent blank

While dilution and direct measurement data show evidence for a constant blank with a variable effect depending on sample size, the blank correction may be simplified if blanks, standards, and unknown samples are the same size (@donahue_isotope-ratio_1990 @roberts_14c_2019). If the fraction modern of the blank is 0.8 modern as estimated above, blank samples will show the most effect of a constant contamination blank with current. Figure 3 shows the fraction modern of measured blanks compared to their mean 12C- ion current. Fraction modern of measured blanks does decrease with increasing current, but the effect is small, less than Fm 0.01 over the range of currents measured. This change with current is roughly equal to the variability between blanks, so the error due to the use of a current-dependent blank based on samples measured over multiple wheels will likely be larger than a current independent blank correction based on blanks measured on each wheel.

LBC eqn

$$Fm_s = Fm_m - Fm_{blk}\frac{Fm_{std} - Fm_m}{Fm_{std}}$$

Data for blanks

Constant current blank is likely the best model, but we simplify to a large blank correction since blanks vary by run, currents are relatively similar between blanks, unknowns and standards, and the simpler model works well within the error of the method.

@@

The difference between reference samples and graphite consensus does not vary systematically by fraction modern, indicating that the method has no significant bias by fraction modern (Figure 6).

@@

## ideas for further development of method

@@

This work is supported by NSF cooperative agreement. We are grateful to C. McIntyre, R. Hansman, G. Salazar, T. Ognibene, T. Broeck.

@@

>>>>

Measurements are corrected for process and measurement blank using a linear correction that assumes that the corrected fraction modern of blank samples should be zero, samples with the fraction modern of the standard need no correction, and applies a proportional correction to samples of intermediate fraction modern. (@roberts_14c_2019) The correction uses measured values of standard gas and blank samples for correction of contaminant carbon and machine blank.  The errors in the blanks and standards are propagated with errors of the samples to produce the final error estimate for the sample. This correction does not account for the effect of a blank with constant mass or current on samples of varying mass or current. For all measurements used to assess the performance of the method and for the Haiti core comparison, sample, standard, and blank masses were kept constant. Sample current did vary, but we believe the linear correction described above produces the most accurate results within this range. We discuss this constant contaminant blank in the results below. Data reduction is done using a package written in R (@r_core_team_r_2022) which is published on github (<https://github.com/blongworth/hgis>).

@@

@@

Initial work with the NEC source established optimal source parameters for consistent source performance. We used tank CO2 connected through a regulator to the ion source to measure source performance over a range of CO2 flows. Best currents for the source are around 10uA 12C- or roughly 1/10 of graphite target current. Best currents were produced at a CO2 flow rate of 7.5uLs-1 CO2, but the source produces near optimal currents with flows from 7.5-25 ul/min CO2. Adding helium to the gas mixture produced similar currents to pure CO2 at the same CO2 flow rates. This is generally comparable with other hybrid sources @Xu2007, although improved target design and continued optimization of source parameters may improve performance @Salazar2013. Typical currents for this study were between 4-9uA and average current for all samples was `r curmean`. Several studies have shown best currents at lower CO2 flow rates (@Xu2007, @Ruff2010). While efficiency of the source is best at very low flow rates (<1.5ul/min), measuring samples with flow set for optimal current produces more precise results due to higher count rates.

Source currents are variable by day and by target under similar source, inlet, and sample conditions. Capillary condition, source and gas delivery arm alignment, slight variability in target geometry, source condition, and performance of the ionizer and cesium may all be factors. Current also varies by target for a given set of gas and source conditions. Capillary clogging can be an issue, especially when sampling from vials with acid.

@@

### Currents and flows

@@

@@

Dilution ratio - 1:1 to 1:3 during analysis

Work with currents and flows (above and figure 2) predicts that optimal currents should be produced until less than 2mL CO2 (in ug?) remains in the sample vial. Tests of a range of sample sizes support this estimate and show that no reduction in measured current is seen in samples larger than XXX mg C which corresponds to XXX mg CaCO3 sample material. To ensure ample sample for optimal measurements, we chose 15 mg as a target sample mass. To ensure that samples started at or slightly above atmospheric pressure, we injected clean helium with a syringe to equilibrate pressure in the vials when the reaction with acid was complete.

## Contaminant carbon

@@

### Current dependent blank

@@

@@

### Current independent blank

While dilution and direct measurement data show evidence for a constant blank with a variable effect depending on sample size, the blank correction may be simplified if blanks, standards, and unknown samples are the same size (@donahue_isotope-ratio_1990 @roberts_14c_2019). If the fraction modern of the blank is 0.8 modern as estimated above, radiocarbon-free samples will show the most effect of a constant contamination blank with current. Figure 3B shows the fraction modern of measured blanks compared to their mean 12C- ion current. Fraction modern of measured blanks does decrease with increasing current, but the effect is small, less than Fm 0.01 over the range of currents measured. This change with current is roughly equal to the variability between blanks, so the error due to the use of a current-dependent blank based on samples measured over multiple wheels will likely be larger than a current independent blank correction based on blanks measured on each wheel. Constant current blank is likely the best model, but we simplify to a large blank correction since blanks vary by run, currents are relatively similar between blanks, unknowns and standards, and the simpler model works well within the error of the method.

We use equation 1 for this correction.

$$ R_{samp} = R_{meas} - \frac{R_{blank}(R_{std}-R_{meas})}{R_{std}} $$

Where $R_{samp}$ is the ratio of the unknown sample, $R_{meas}$ is the measured ratio, $R_{blank}$ is the ratio of the blank, and $R_{std}$ is the ratio of the normalizing standard.

@@

The difference between reference samples and graphite consensus does not vary systematically by fraction modern, indicating that the method has no significant bias by fraction modern (Figure 4).

@@

ideas for further development of method

@@

This work is supported by NSF cooperative agreement. We are grateful to C. McIntyre, R. Hansman, G. Salazar, T. Ognibene, T. Broek.

@@

# References
